---
title: "Estimating Likelihood"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# Excrcise: claculate tree likelihood

Try coding your own Likelihood algorithm that calculates likelihood of a state pattern given: (1) tree, (2) rate matrix with specified parameters, and (3) specified initial vector of probabilities at tree root. Make this algorithm working for a 3-taxon tree. Your toy algorithm should work like the Felsentein's pruning algorithm from today's lecture. Use binary characters for brevity.


Before, coding the Ln algorithm, simulate a tree and note the main data structures that store the tree topology and branch lengths

```{r}
library(ape)
library(phytools)

set.seed(100)
tr<-pbtree(n=3, scale=1, b=1, d=0)
plot(tr)
nodelabels()
edgelabels()
tiplabels()

# save tree file, this is optional
#write.tree(tr, file='3_taxon.tree')
```

You will need these data objects for your Ln algoritm *tr\$edge*, *tr\$edge.length*
```{r}
# tree topology
tr$edge

# branch lengths
tr$edge.length

```



# Answer to the Exercise: Calculating Likelihood on tree
See the slides for Felsentein's pruning algorithm.

The general steps of the algorithm are as follows:

1. Initialize $Q$ matrix and $\pi$ (initial vector at the tree root)
2. Recode data at tips as probabilities
3. Calculate conditional likelihoods $L_c$ at each node by traversing the tree from tips to roots. 
    * For two branches -- left and right ($b_l$ and $b_r$) -- and their corresponding tip probabilities ($T_l$, $T_r$, both column vectors), the conditional likelihood is  $L_c=e^{Qb_l}T_l*e^{Qb_r}T_r$. 
    * Since R does not have inbuilt functionality for column vectors, $T_l$, $T_r$ can be treated as row vectors through the following conversion $L_c=T_l(e^{Qb_l})^\textbf{T}*T_r(e^{Qb_r})^\textbf{T}$, where $^\textbf{T}$ is a matrix transpose
4. Upon reaching root, the tree likelihood is $L=sum(L_c\pi)$


#### 1. Initializing Q and pi
```{r}

# install packages exmp if you do not have it
library (expm)

Q <- matrix(
  c(
    -0.1, 0.1,
    0.1, -0.1
    ), 2,2, byrow = T)

pi=c(0.5, 0.5)
Q
pi

```



#### 2. Recoding data at tips as probabilities
```{r}
# read in tree, this is optional if the tree was not simulated
# tr <- read.tree(file='3_taxon.tree')

# Let's imagine that our taxa at tips (t1, t2, t3) have the following character states (we use a binary character)
data <- c(1,0,1)
names(data) <- c('t1', 't2', 't3')
data
```

We represent the tip probability vectors as matrix
```{r}
data.probs <- matrix(c(
  0,1,
  1,0,
  0,1
  ), 3,2,byrow=T
)
rownames(data.probs) <- c('t1', 't2', 't3')
data.probs
```

#### 3. Calculating conditional likelihoods
```{r}
# Let's plot a tree to see its structure that is encoded on the objects tr$edge and tr$edge.length
plot(tr)
nodelabels()
edgelabels()
tiplabels()

# tree topology
tr$edge
# branch lengths
tr$edge.length
```


First we calculate Lc for node 5 with tips t3 and t2
```{r}
# get ids of two branches node5 -> t3 and node5 -> t2 in tr$edge object (they are 3 and 2)
t3t2.ids=which(tr$edge[,1]==5)

# get branch lengths for those branches
br_l_r=tr$edge.length[t3t2.ids]

# calculate conditional likelihood using matrix exponential expm(), and matrix multiplication %*%
L5=data.probs[3,] %*% t(expm(Q*br_l_r[1]))*(data.probs[2,]%*%t(expm(Q*br_l_r[2])))
L5
```

Now we calculate root likelihood (node4)
```{r}
# get ids of two branches node4 -> t1 and node4 -> node5 
n5t1.ids=which(tr$edge[,1]==4)
# get branch lengths for those branches
br_l_r=tr$edge.length[n5t1.ids]
br_l_r

# calculate conditional likelihood at root using matrix exponential expm(), and matrix multiplication %*%
Lr=(data.probs[1,]%*%t(expm(Q*br_l_r[1])))*(L5 %*% t(expm(Q*br_l_r[2])))
Lr
```

#### 4. Calculating tree likelihood
```{r}
L=sum(Lr*pi)

# likelihood and Log likelihood of the tree
L
log(L)
```



