---
title: "Ex1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Welcome to the ForBio Course wiki for modeing discrete traits!

To run the exercise you need to have [R](https://www.r-project.org/) and [RStudio](https://www.rstudio.com/) preinstalled.
Also, instal the required packages by running the following code from within Rstudio:

```{r}
#install.packages(c("corHMM", "phytools")
library("corHMM")
library("phytools")
```


# Exercise 1. Model Selection and Ancestral State Reconstruction

**Aim:** Reconstruct *ancestral states* (ANS) of character evolution. Compare models using 
statistical modeling selection criteria (AIC).

**Instruction:**
1. Simulate tree.
2. Simulate character evolution on the tree under $M_0$ (asymmetric model).
3. Fit two models to data ($M_0$: asymmetric and $M_1$: symmetric). Reconstruct ancestral states.
4. Compare models using AIC.



## 1. Simulate tree
Simulate tree using pure birth process

```{r}
tree<-pbtree(n=200, scale=100, b=1, d=0)
plot(tree)
```



## 2. Simulate character evolution on the tree under asymmetric model.

For simulation character evolution you need a tree and rate matrix $Q$.

```{r}
# make rate matrix Q
Q <- matrix(
  c(
    -0.03, 0.03,
    0.1, -0.1
    ), 2,2, byrow = T)

Q

```

Simulate character evolution on tree using Q and plot it.
```{r}
hist <- sim.history(tree, Q, nsim=1)
plot(hist)
```

## 3. Fit two models to data. Reconstruct ancestral states. 

For running inference one needs a tree, pre-specified parametrization of rate matrix, and data (=characyter). The inference returns rates
of character evolution ($Q$) and character state probabilities at tree nodes.

## Inference using Symmetrical Model $M_1$
```{r}
# one-rate symmetrical model
Q.Sym <- matrix(
  c(
    NA, 1,
    1, NA
  ), 2,2, byrow = T)

Q.Sym

# Inference
taxa <- cbind(hist$tip.label, hist$states)
Recon_Q.Sim <- rayDISC(hist, taxa, rate.mat=Q.Sym, node.states="marginal", 
        model="ARD", root.p="maddfitz")

# infered rate matrix
Recon_Q.Sim



```

Plot ANS
```{r}
plotRECON(tree, Recon_Q.Sim$states, piecolors=c('black', 'red'), title="1-rate Model")
```


##  Inference using Asymmetrical Model ($M_0$)

```{r}
# 2-rate symmetrical model
Q.Asym <- matrix(
  c(
    NA, 1,
    2, NA
  ), 2,2, byrow = T)

Q.Asym

# Inference
taxa <- cbind(hist$tip.label, hist$states)
Recon_Q.Asim <- rayDISC(hist, taxa, rate.mat=Q.Asym, node.states="marginal", 
                       model="ARD", root.p="maddfitz")

# infered rate matrix
Recon_Q.Asim

# plot ANS
plotRECON(tree, Recon_Q.Asim$states, piecolors=c('black', 'red'), title="1-rate Model")

```



##   4. Compare models using AIC.  

```{r}
# this objects contain all necessary info
Recon_Q.Sim
Recon_Q.Asim

# AIC is here
Recon_Q.Sim$AIC

# delta AIC
Recon_Q.Sim$AIC-Recon_Q.Asim$AIC
```


##   Answer the Questions 

1. Which model performs better, $M_0$ or $M_1$?
2. Does model choice affect the reconstructed state probabilities? How we can assess it?
3. Run the same simulations with varying number of tips, for example, pbtree(n=10, ...), pbtree(n=500, ...), .... 
How does number of data points affect inference?


