---
title: "Ex 2"
output: html_notebook
---


# Exercise 2. Testing Correlated Character Evolution


**Aim:** Test if character evolution is (un)correlated using AIC.

**Instruction:**
1. Simulate tree.
2. Simulate character evolution on the tree under the correlated model $M_0$.
3. Fit two models to data ($M_0$: correlated and $M_1$: uncorrelated).
4. Compare models using AIC.

For this exercise you need to source some precooked functions from the file `SMM_functions.R`; they are used for constructing amalgamated reate matrices. Copy this file from GitHub to you working directory and execute the code below.

```{r}
# useful functions 
source("SMM_functions.R")
```



## 1. Simulate tree.

Let's make a correlated character where tail color evolution is correlated with tail shape (long, short)

```{r}
# initialize shape character
char.state<-c("L", "S")
rate.param<-c(0.1, 0.1)
TL<-init_char_matrix(char.state, rate.param, diag.as=0)
diag(TL)<-rate.param*-1
TL

# initialize color character
char.state<-c("r", "b")
rate.param<-c(0.1, 0.1)
COL<-init_char_matrix(char.state, rate.param, diag.as=0)
diag(COL)<-rate.param*-1
COL

# amalgamate two characters
Q.ind<-comb2matrices(TL, COL, controlling.state=2, name.sep="", diag.as="")
Q.ind

# make the character correlated
Q.cor <- Q.ind
Q.cor[2,4] <- 0.3
Q.cor[3,4] <- 0.3
diag(Q.cor) <- 0
diag(Q.cor) <- -rowSums(Q.cor)
Q.cor

```



## 2. Simulate character evolution on the tree under the correlated model $M_0$.

```{r}
tree<-pbtree(n=100, scale=200)
sim.h<-sim.history(tree, Q.cor, nsim=1, anc=setNames(c(1, 0, 0, 0), colnames(Q.cor) ) )
plot(sim.h)
```



                                                          
## 3. Fit two models to data ($M_0$: correlated and $M_1$: uncorrelated).

Let's first construct rate matrices that we will use for the ML inference
```{r}
# matrix template for inference
char.state<-c("T", "C")
rate.param<-c(1, 1)
TL<-init_char_matrix(char.state, rate.param, diag.as=0)

char.state<-c("r", "b")
rate.param<-c(1, 1)
COL<-init_char_matrix(char.state, rate.param, diag.as=0)

# uncorrelated matrix
M.equal<-comb2matrices(TL, COL,  name.sep="", diag.as="", non.rate.as=NA)
M.equal

# correlated matrix
M.corr <- M.equal
M.corr[2,4] <- 2
M.corr[3,4] <- 2
M.corr
```


### ML Inference

                          ####

```{r}
# uncorrelated matrix
taxa <- cbind(hist$tip.label, sim.h$states)
out.equal <- rayDISC(sim.h, taxa, rate.mat=M.equal, node.states="marginal", model="ARD", root.p="maddfitz")

# correlated matrix
out.corr <- rayDISC(sim.h, taxa, rate.mat=M.corr, node.states="marginal", model="ARD", root.p="maddfitz")
```



## 4. Compare models using AIC.

```{r}
# our results
out.equal
out.corr 

# delta AIC
out.equal$AIC-out.corr$AIC
```


##   Answer the Questions 

1. Which model is suggested by AIC: the uncorrelated or correlated one?
2. What are the differences inn Q between correlated and uncorrelated models?
3. Do you know any real life cases where evolution of discrete traits might be correlated?


